---

title: Baseball player salary prediction
author: Roxana Virlan
output: html_notebook
date: 3/12/2020
version: 1
---


```{r}
# Packages needed for this coursework are : tidyverse, ggplot2, validate 
library(tidyverse)
library(ggplot2)
library(validate)



```


# 1. Organise and clean the data

## 1.1 Subset the data into the specific dataset allocated


```{r}
my.2teams.df <- subset(CS5801.data, teamID.x=="CHN" | teamID.x=="LAN")
summary(my.2teams.df)
str(my.2teams.df)

```


## 1.2 Data quality analysis
 
*Provide a description of a comprehensive plan to assess the quality of the data, and document your findings.  Include all variables/columns (5 marks) from the data set and provide a full implementation (5 marks).  NB even if no data quality issues are identified you should still check and report.*
```{r}
# To check for quality issues I will use the package validate and set some rules for checking for any issues.
1. Weight of the players should be between 150-300
2. Height of the players should be between 66-73.2
3. PlayerID should not be any duplicated.
4. Should not be any missing values
5. Age and birthDate should match.
6  Must not be any negative values.
```


```{r}
my.2teams.df.rules<-validator(okteamID.x=is.element(teamID.x,c                                    ("CHN","LAN")),
                              NonNegGame = Game >= 0,
                         noduplicates=is_unique(playerID),
                         agevalue=age~birthDate) 
qual.check <- confront(my.2teams.df,my.2teams.df.rules)
summary(qual.check)
```

```{r}
barplot(qual.check, xlab = "")
```

 
 

#Observed from the summary output that the playerID was a character and two of them were repeating decided to change the entire column to factor levels. "To get a summary of a variable, it needs to be a factor."(Zumel,Mount,Practical Data Science with R, 2014  p37)
```{r}
my.2teams.df$playerID =as.factor(my.2teams.df$playerID)
str(my.2teams.df)
```




```{r}

```


```{r}
my.2teams.df$age <- as.integer(my.2teams.df$age)
print(my.2teams.df$age)
```


 
## 1.3 Data cleaning  
 
*Explain any data quality issues found in 1.2 (5 marks), justify and document the responses made (if any) (5 marks).*
```{r}
#Looking at the game data I could see that the min value is -38 and if this data shows the number games played it cannot be a minus value. So I will change it to positive 
```
```{r}
my.2teams.df$G[my.2teams.df$G ==-38]=38
summary(my.2teams.df)

```
# Next cleaning is checking to see if there is any issues with with date and age.
From using the validate package a fail between age and birthdate could be observed. I decided to remove the entire column of birthdate as I could see more mistakes in it.


```{r}
my.2teams.df<- subset (my.2teams.df, select = -birthDate)
my.2teams.df
#https://www.tutorialspoint.com/how-to-remove-a-column-from-an-r-data-frame
```

```{r}
#removing duplicate row
 my.2teams.df=my.2teams.df[-c(25,71),]
#(https://sites.ualberta.ca/~ahamann/teaching/renr690/R_Cheat_Data.pdf)
summary(my.2teams.df)
```


```{r}

```

```{r}
#Cheking on the website wearefanatics.com/baseball-players-shapes-sizes/ I could see the values that the weight(151-300lb) and height(66inch-73.2) should be in this range.Remove the outliers.
```
```{r}

```

```{r}
is.na (my.2teams.df)
str(is.na(my.2teams.df))
# did check for missing values in the data and could not find any missing values.
```


# 2. Exploratory Data Analysis (EDA)

## 2.1 EDA plan
# Answering the folowing questions
# What data types am I dealing with?

What uncertainty surrounds this estimate of typicality?
What type of variation occurs within my variables?
Does the data set contain outliers?
What type of covariation occurs between my variables?
More specifically, if there are any factors, do they have an effect?

*Outline a suitable plan to explore, describe and visualise your data. (5 marks)* 
In order to explore , describe and visualise the data is going to be explored with the help of summary function and every variable is going to be explored alone and in pairs in order to see any relationships with the help of visual aids like histograms, boxplots,scatterplots.
## 2.2 EDA and summary of results  

*Undertake and summarise the findings of your data exploration, particularly with respect to the research questions.  Use appropriate summary statistics (uni- and multi-variate) and visualisations. (10 marks)*
```{r}
library(ggplot2)
#Distribution of variables
ggplot(data = my.2teams.df, aes(x=G)) + geom_histogram(bins = 20) + theme_classic()
ggplot(data = my.2teams.df, aes(x=R)) + geom_histogram(bins = 20) + theme_classic()
ggplot(data = my.2teams.df, aes(x=H)) + geom_histogram(bins = 20) + theme_classic()
ggplot(data = my.2teams.df, aes(x=AB)) + geom_histogram(bins = 20) + theme_classic()
ggplot(data = my.2teams.df, aes(x=RBI)) + geom_histogram(bins = 20) + theme_classic()
ggplot(data = my.2teams.df, aes(x=weight)) + geom_histogram(bins = 20) + theme_classic()
ggplot(data = my.2teams.df, aes(x=height)) + geom_histogram(bins = 20) + theme_classic()
ggplot(data = my.2teams.df, aes(x=salary)) + geom_histogram(bins = 20) + theme_classic()
ggplot(data = my.2teams.df, aes(x=career.length)) + geom_histogram(bins = 20) + theme_classic()
ggplot(data = my.2teams.df, aes(x=age)) + geom_histogram(bins = 20) + theme_classic()

```
#The visualisation of the histograms shows that most of the data is skewed to the right and that outliers are present.
#Looking at the histogram of weight I could see an outlier which is 485 pounds.
#creating scatterplots to check for any relatshionships
```{r}
#I will use boxplots to check for outliers
boxplot(my.2teams.df$G, main="Boxplot of game")
boxplot(my.2teams.df$R, main="Boxplot of Runs")
boxplot(my.2teams.df$H, main="Boxplot of Hits")
boxplot(my.2teams.df$AB, main="Boxplot of At Bats")
boxplot(my.2teams.df$RBI, main="Boxplot of Runs Batted In")
boxplot(my.2teams.df$weight, main="Boxplot of Weight")
boxplot(my.2teams.df$height, main="Boxplot of Height")
boxplot(my.2teams.df$salary, main="Boxplot of Salary")
boxplot(my.2teams.df$career.length, main="Boxplot of Career length")
boxplot(my.2teams.df$age, main="Boxplot of age")
```
#Looking at these boxplots I could see that within the dataset there are outliers and I will use the 1.5times  formula to remove them:"an outlier is a value more than 1.5 times the interquartile range above the 3rd
quartile, or below the 1st quartile".(Crawley, p80)  I will apply this rule only to weight and height.
```{r}
Q1 <- quantile(my.2teams.df$weight, .25)
Q3 <- quantile(my.2teams.df$weight, .75)
IQR <-IQR(my.2teams.df$weight)
my.2teams.df <- subset(my.2teams.df, my.2teams.df$weight >(Q1 - 1.5*IQR) & my.2teams.df$weight< (Q3 + 1.5*IQR))
dim(my.2teams.df)
 #https://www.statology.org/remove-outliers-r/

```


```{r}
Q1 <- quantile(my.2teams.df$height, .25)
Q3 <- quantile(my.2teams.df$height, .75)
IQR <-IQR(my.2teams.df$height)
my.2teams.df <- subset(my.2teams.df, my.2teams.df$height >(Q1 - 1.5*IQR) & my.2teams.df$height< (Q3 + 1.5*IQR))
dim(my.2teams.df)
```
```{r}
#For better visualisation of salary I will use a density plot to check the overall shape of the curve.
ggplot(my.2teams.df)+geom_density(aes(x=salary)) 
```
# As the salary is heavily skewed to the right I will use log transformation to better visulise data.
```{r}
ggplot(my.2teams.df)+geom_density(aes(x=salary)) +        scale_x_log10(breaks=c(100,1000,10000,100000))+annotation_logticks (sides="bt")
(Zumel N. and Mount J,2014)
```
```{r}
boxplot(my.2teams.df$G, main="Boxplot of game")
boxplot(my.2teams.df$R, main="Boxplot of Runs")
boxplot(my.2teams.df$H, main="Boxplot of Hits")
boxplot(my.2teams.df$AB, main ="Boxplot of At Bats")
boxplot(my.2teams.df$RBI, main="Boxplot of Runs Batted In")
boxplot(my.2teams.df$weight, main="Boxplot of Weight")
boxplot(my.2teams.df$height, main="Boxplot of Height")
boxplot(my.2teams.df$salary, main="Boxplot of Salary")
boxplot(my.2teams.df$career.length, main="Boxplot of Career length")
boxplot(my.2teams.df$age, main="Boxplot of age")
```





```{r}
ggplot(data = my.2teams.df, aes(x=salary, y=G)) + geom_point() + theme_classic()
ggplot(data = my.2teams.df, aes(x=salary, y=R)) + geom_point()  + theme_classic()
ggplot(data = my.2teams.df, aes(x=salary, y=H)) + geom_point()  + theme_classic()
ggplot(data = my.2teams.df, aes(x=salary, y=AB)) + geom_point()  + theme_classic()
ggplot(data = my.2teams.df, aes(x=weight, y=height)) + geom_point()  + theme_classic()
ggplot(data = my.2teams.df, aes(x=salary, y=career.length)) + geom_point() + theme_classic()
ggplot(data = my.2teams.df, aes(x=salary, y=age)) + geom_point()  + theme_classic()
```
# The collection of scatterplots do not show that any of the variables is clearly liniar, but some show a liniar trend.

```{r}
# Focusing only on the continuous explanatory variables - check their correlations with the Salary. We want to do this only for the continuous variables, so can look to remove the column that is binary from this plot. (This is done so that the pairs plot is legible and that we can run a corr function on the resulting dataframe)
baseball<-subset(my.2teams.df,select=c("salary","G","R","H","AB","weight","height","career.length","age"))
pairs(baseball,panel = panel.smooth)
cor(baseball)
```




## 2.3 Additional insights and issues

*Highlight potential further issues or insights uncovered in 2.2.  This might include follow up to findings from your initial EDA.  We accept that the boundary between 2.2 and 2.3 is somewhat arbitrary so use your judgement and maximise good structure and readability. (5 marks)*
#During EDA I did use histograms to check for skewness and it did show that most of them are skewed to the right except weight and height.
I used boxplots to visualize for outliers and I have used the *1.5 formula on the data set to remove outliers. And I removed the rows that showed outliers remaining in the data sets with 73 rows and 14 variables



# 3. Modelling

## 3.1 Build a model for player salary

*Given the research question (i.e., the target attribute of salary) outline an analysis plan that incorporates/references any findings from the data cleaning (1.3) and EDA (2.2)  (5 marks). Use R to build a suitable model (10 marks).*  
*NB Submissions where suitable models do not have good fit due to the nature of the data will not be penalised.*  


```{r}
fit<-lm(salary~G+R+H+AB+RBI+weight+height+career.length+age, data=my.2teams.df)
summary(fit)
```

                 
```
## 3.2 Critique model using relevant diagnostics
```{r}
plot(fit)
```
#To determine if the model fulfils the regression assumptions the plot(fit) was used.
The residuals and predicted values look random.
The QQ plot shows that most of the points fall on the reference line.
The scale-location plot looks random with no patterns and the homoscedasticity assumptions is not violated.
The last plot should show presence of any outliers.(Nikita E, 2020)
```{r}
library(car)
vif (fit)
```
#From testing the multicollinearity assumptions the Vif scores are bigger then 10 : my.2teams.df$H, my.2teams.df$RBI, my.2teams.df$R, my.2teams.df$AB(Nikita E, 2020) 
```{r}

```


*Offer an interpretation of the model characteristics, goodness of fit and graphical diagnostics (5 marks) for the model built in 3.1. Explain any potential weaknesses (5 marks).*
#The adjusted r squared=0.34, the higher the value the better fit of the model
#The p value of the model=2.837e-05, which is statistically significant

## 3.3 Suggest improvements to your model

*Based on the findings in 3.2 articulates possible alternative approaches to address them (5 marks).*
#I will use the log transformation for salary and use step function to simplify the model
```{r}
lm(log(salary)~G+R+H+AB+RBI+weight+height+career.length+age,data=my.2teams.df)
summary(lm)
```



# 4. Extension work

## 4.1 Model the likelihood of a player having scored a Hit (using the hit.ind variable provided).

*Given this second research question (i.e., involving the binary target attribute) provide a plan of analysis based on relevant EDA for this attribute (10 marks). The model is described, explained and critiqued (10 marks).*
*NB Submissions where suitable models do not have good fit due to the nature of the data will not be penalised.* 


```{r}
model<-glm(hit.ind~)
```

# References  

*Add any references here. NB You can either do this manually or automatically with a `.bib` file (which then must be submitted along with your `.Rmd` file).  See the RMarkdown [documentation](https://bookdown.org/yihui/rmarkdown-cookbook/bibliography.html) for guidance.*    
#https://williamkoehrsen.medium.com/data-analysis-with-python-19434f5d6324
https://sites.ualberta.ca/~ahamann/teaching/renr690/R_Cheat_Data.pdf
https://www.tutorialspoint.com/how-to-remove-a-column-from-an-r-data-frame
https://www.statology.org/remove-outliers-r/
Crawley()Statistics: an introduction using R
Zumel N. and Mount J.,(2014) Practical Data Science with R,Manning Publications Co. 
Nikita E.(2020)Introduction to statistics using R